### 2015-08-07 background.tex
some fixes

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break; // racy
    ...
  }
  if (offset < 0) return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Racy entry point in the generic\_nvram Linux 4.0 driver}
\label{fig:data_race_example}
%\vspace{-2mm}
\end{figure}

@@

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break;
    ...
  }
  if (offset < 0) {
    mutex_unlock(&nvram_mutex); // unlock
    return -EINVAL;
  }
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Introducing a lock to eliminate the data races in the previous example}
\label{fig:lock_example}
%\vspace{-2mm}
\end{figure}

@@

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program. Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks initially. In addition, every $\mathit{LS}_s$ is initialized to the set of all locks manipulated by the program since initially each access to $s$ is (vacuously) protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
%
After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
%
When $T$ accesses a shared variable $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
%
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
%\vspace{-5mm}
\caption{Applying lockset analysis on a concurrent program}
\label{fig:locksets}
%\vspace{-3mm}
\end{figure}

>>>>

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break; // racy
    ...
  }
  if (offset < 0) return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\vspace{-2mm}
\caption{Racy entry point in the generic\_nvram Linux 4.0 driver.}
\label{fig:data_race_example}
%\vspace{-2mm}
\end{figure}

@@

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break;
    ...
  }
  if (offset < 0) {
    mutex_unlock(&nvram_mutex); // unlock
    return -EINVAL;
  }
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\vspace{-2mm}
\caption{Introducing a lock to eliminate the races in the previous example.}
\label{fig:lock_example}
%\vspace{-2mm}
\end{figure}

@@

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program. Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks on program start. In addition, every $\mathit{LS}_s$ is initialized to the set of all locks manipulated by the program since initially each access to $s$ is (vacuously) protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
%
After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
%
When $T$ accesses $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
%
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
%\vspace{-5mm}
\caption{Applying lockset analysis on a concurrent program.}
\label{fig:locksets}
%\vspace{-3mm}
\end{figure}

### 2015-08-07 whoop.tex
some fixes

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented our analysis in a practical tool, \whoop, that can be applied directly to driver source code.
%
\COULDCUT{Our experimental evaluation (\S\ref{evaluation}) demonstrates that \whoop has value as a stand-alone analyzer, and that results from our analysis can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.}

@@

$\mutexunlock{m};$ & mutex unlock, where $m \in \mutexes$\\
\bottomrule
\end{tabular}
\caption{The allowed statements in our simple programming model}
\label{fig:statements}
%\vspace{-2mm}
\end{figure}

\noindent\textbf{Semantics }
%
Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.  The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{thread state}.  Each thread state has an associated thread template $T$, the type of the thread (and multiple threads may have the same associated template), and consistes of an id (drawn from $\ids$), an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables $\variables{T}$.  If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.

@@

$\mutexunlock{m};$ & $\currentlockset{i} = \currentlockset{i} \setminus \{ m \};$ \\
\bottomrule
\end{tabular}
\caption{Instrumenting statements for lockset analysis}
\label{fig:instrumentation}
%\vspace{-2mm}
\end{figure}

@@

\end{tabular}
\caption{The sequential program to be analyzed to prove race-freedom for a pair of thread templates}
\label{fig:sequentialization}
%\vspace{-2mm}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=.99\linewidth]{img/whoop.pdf}
%\vspace{-2mm}
\caption{The \whoop architecture, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
%\vspace{-3mm}
\label{fig:whoop}
\end{figure*}

@@

\begin{figure}[t]
\begin{lstlisting}
const struct file_operations nvram_fops = {
  .llseek = nvram_llseek,
  .read = read_nvram,
  .write = write_nvram,
  .unlocked_ioctl = nvram_unlocked_ioctl
};
\end{lstlisting}
%\vspace{-2mm}
\caption{Entry point definitions in the generic\_nvram driver}
\label{fig:entrypoints}
%\vspace{-2mm}
\end{figure}

@@

\noindent\textbf{Generating Loop and Procedure Summaries }
%
Early in the development of \whoop, we experimented with analyzing recursion-free drivers using full inlining.  We found that this did not scale to large drivers, and some drivers do exhibit recursion, e.g.\ the r8169 ethernet driver (see \S\ref{evaluation}).

>>>>

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented our analysis in a practical tool, \whoop, that can be applied directly to driver source code.

@@

$\mutexunlock{m};$ & mutex unlock, where $m \in \mutexes$\\
\bottomrule
\end{tabular}
\caption{The allowed statements in our simple programming model.}
\label{fig:statements}
%\vspace{-2mm}
\end{figure}

\noindent\textbf{Semantics }
%
Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.  The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{thread states}.  Each thread state has an associated thread template $T$, the type of the thread (and multiple threads may have the same associated template), and consists of an id (drawn from $\ids$), an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables $\variables{T}$.  If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.

@@

$\mutexunlock{m};$ & $\currentlockset{i} = \currentlockset{i} \setminus \{ m \};$ \\
\bottomrule
\end{tabular}
\caption{Instrumenting statements for lockset analysis.}
\label{fig:instrumentation}
%\vspace{-2mm}
\end{figure}

@@

\end{tabular}
\caption{The sequential program to be analyzed to prove race-freedom for a pair of thread templates.}
\label{fig:sequentialization}
%\vspace{-2mm}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=.99\linewidth]{img/whoop.pdf}
%\vspace{-2mm}
\caption{The \whoop architecture, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools.}
%\vspace{-3mm}
\label{fig:whoop}
\end{figure*}

@@

\begin{figure}[t]
\begin{lstlisting}
const struct file_operations nvram_fops = {
  .llseek = nvram_llseek,
  .read = read_nvram,
  .write = write_nvram,
  .unlocked_ioctl = nvram_unlocked_ioctl
};
\end{lstlisting}
\vspace{-2mm}
\caption{Entry point definitions in the generic\_nvram driver.}
\label{fig:entrypoints}
%\vspace{-2mm}
\end{figure}

@@

\noindent\textbf{Generating Loop and Procedure Summaries }
%
Early in the development of \whoop, we experimented with analyzing recursion-free drivers using full inlining.  We found that this did not scale to large drivers, and that some drivers do exhibit recursion, e.g.\ the r8169 ethernet driver (see \S\ref{evaluation}).
